Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-03-01 paper/main.tex
update references to supp figs to match new order
minor rephrasings to clarify intent of BERT/LDA comparisons

% supplementary tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}
% supplementary figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\forcesCorrs}{3}
\newcommand{\bosCorrs}{4}
\newcommand{\jaccard}{5}
\newcommand{\ldaVsBERT}{6}
\newcommand{\individualKnowledgeMapsA}{7}
\newcommand{\individualKnowledgeMapsB}{8}
\newcommand{\individualKnowledgeMapsC}{9}
\newcommand{\individualLearningMapsA}{10}
\newcommand{\individualLearningMapsB}{11}
% supplementary results
\newcommand{\suppResults}{\textit{Supplementary results}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag of words'' text embedding model~\citep[LDA;
][]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23},
can leverage additional textual information such as complex grammatical and 
semantic relationships between words, higher-order syntactic structures, 
stylistic features, and more. We considered using transformer-based models in 
our study, but we found that the text embeddings derived from these models were 
surprisingly uninformative with respect to differentiating or otherwise 
characterizing the conceptual content of the lectures and questions we used 
(see \suppResults). We suspect that this reflects a broader challenge in 
constructing models that are both high-resolution within a given domain (e.g., 
the domain of physics lectures and questions) \textit{and} sufficiently broad 
as to enable them to cover a wide range of domains. Essentially, these 
``larger'' language models learn these more complex features of language through 
pre-training on enormous and diverse text corpora. But as a result, their 
embedding spaces also ``span'' an enormous and diverse range of conceptual 
content, sacrificing a degree of specificity in their capacities to distinguish 
subtle conceptual differences within a more narrow range of content. 
In comparing our LDA model (trained specifically on the lectures used in our 
study) to a larger transformer-based model (BERT), we found that LDA provides 
both coverage of the requisite material and specificity at the level of 
individual questions, while BERT essentially relegates the contents of both 
lectures and all quiz questions (which are all broadly about ``physics'') to a 
tiny region of its embedding space, thereby blurring out meaningful distinctions 
between different specific concepts covered by the lectures and questions 
(Supp.~Fig.~\ldaVsBERT). We note that these are not criticisms of BERT, nor of 
other large language models trained on large and diverse corpora. Rather, our 
point is that simpler models trained on relatively small but specialized 
corpora can outperform much more complex models trained on much larger corpora 
when we are specifically interested in capturing subtle conceptual differences 
at the level of a single course lecture or question. On the other hand, if our 
goal had been to choose a model that generalized to many different content 
domains, we would expect our LDA model to perform comparatively poorly to BERT 
or other much larger models. We suggest that bridging this tradeoff will be
an important challenge for future work.

>>>>

% supplementary tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}
% supplementary figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\forcesCorrs}{3}
\newcommand{\bosCorrs}{4}
\newcommand{\individualKnowledgeMapsA}{5}
\newcommand{\individualKnowledgeMapsB}{6}
\newcommand{\individualKnowledgeMapsC}{7}
\newcommand{\individualLearningMapsA}{8}
\newcommand{\individualLearningMapsB}{9}
\newcommand{\jaccard}{10}
\newcommand{\ldaVsBERT}{11}
% supplementary results
\newcommand{\suppResults}{\textit{Supplementary results}}

@@

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag of words'' text embedding model~\citep[LDA;
][]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23},
can leverage additional textual information such as complex grammatical and 
semantic relationships between words, higher-order syntactic structures, 
stylistic features, and more. We considered using transformer-based models in 
our study, but we found that the text embeddings derived from these models were 
surprisingly uninformative with respect to differentiating or otherwise 
characterizing the conceptual content of the lectures and questions we used 
(see \suppResults). We suspect that this reflects a broader challenge in 
constructing models that are both high-resolution within a given domain (e.g., 
the domain of physics lectures and questions) \textit{and} sufficiently broad 
as to enable them to cover a wide range of domains. Essentially, these 
``larger'' language models learn these more complex features of language through 
training on enormous and diverse text corpora. But as a result, their 
embedding spaces also ``span'' an enormous and diverse range of conceptual 
content, sacrificing a degree of specificity in their capacities to distinguish 
subtle conceptual differences within a more narrow range of content. 
In comparing our LDA model (trained specifically on the lectures used in our 
study) to a larger transformer-based model (BERT), we found that our LDA model provides 
both coverage of the requisite material and specificity at the level of 
individual questions, while BERT essentially relegates the contents of both 
lectures and all quiz questions (which are all broadly about ``physics'') to a 
tiny region of its embedding space, thereby blurring out meaningful distinctions 
between different specific concepts covered by the lectures and questions 
(Supp.~Fig.~\ldaVsBERT). We note that these are not criticisms of BERT, nor of 
other large language models trained on large and diverse corpora. Rather, our 
point is that simpler models trained on relatively small but specialized 
corpora can outperform much more complex models trained on much larger corpora 
when we are specifically interested in capturing subtle conceptual differences 
at the level of a single course lecture or quiz question. On the other hand, if 
our goal had been to choose a model that generalized to many different content 
areas, we would expect our LDA model to perform comparatively poorly to BERT or 
other much larger general-purpose models. We suggest that bridging this tradeoff 
between high resolution within a single content area and the ability to 
generalize to many diverse content areas will be an important challenge for 
future work.
